Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-16 latex/main.tex
Firt draft of Zarr section

@@

We downloaded 4,484,157 Viridian sequences obtained
from samples collected from January, 2020 to March, 2023 (``batch 1'').
We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using MAFFT v7.525 \cite{Katoh2013},
with the option ‘--keeplength’ to retain the reference coordinate system, discarding insertions.
We converted the MAFFT alignments and all sample metadata to VCF Zarr format \cite{Czech2025},
which are available for download on Zenodo (URL).

@@

Before characterizing recombination events in detail,
we sought to remove probable artifacts.
Systematic errors in SARS-CoV-2 sequences
associated with sequencing protocols \cite{Turakhia2020}
can complicate recombination detection \cite{DeMaio2024}.
% Explain to readers what convincing genomic evidence looks like.
By inspecting copying patterns at sites that differ between recombinant parents,
we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.

Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
illustrate the evidence for three recombination events identified as artifactual.
Explain the three bad patterns:

Compared to a marginally robust one.

Artifactual recombination events can be characterized
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.

When counting such loci, we considered nearly adjacent sites
(separated by at most three bases) as single loci,
to avoid over-counting support from correlated loci,
which could lead to artificial signals of recombination.
Also, when counting supporting loci for a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
Evidence that our simple filter effectively flagged artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events
that we identify as artifactual.
This primer scheme was used to sequence XX of the samples
associated with an artifactual recombination event,
in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
strongly implicating it as a common source of problematic data.

@@

This computation is a good example of type of integrated large-scale data analysis 
that can be easily performed using the combination of tskit and VCF Zarr.
The entire analysis is performed in a short Jupyter notebook, with the 
core calculation requiring about 30 lines of Python code. Retrieving 
data from the base alignments and ARGs for 2.48 million samples 
at the same time over all 27,507 sites took only 6 minutes, with a peak memory
usage of about 5GiB of RAM on a standard desktop computer.

>>>>

\subsubsection*{Pandemic scale alignment storage}
To provide the efficient access to alignment data required for inference with sc2ts, we
developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
Briefly, Zarr is a data storage format designed for scientific data applications
originally developed to store genetic variation data for the
\textit{Anopheles gambiae}
1000 Genomes Project~\cite{Anopheles2017genetic}, and
is now seeing adoption across a wide range of 
applications~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
Zarr is essentially a simple mechanism for storing n-dimensional array
data as a regular grid of compressed chunks, which provides 
high levels of data compression, efficient access across multiple
dimensions, and is ideally suited to modern cloud-based
deployments~\cite{Czech2025}.

We converted the Viridian v05 dataset\cite{Hunt2024}, consisting of 4,484,157 SARS-CoV-2
consensus sequences (125GiB over 48 FASTA files) and 30 metadata fields (1.4
GiB TSV file) to VCF Zarr using a 
reproducible Snakemake\cite{Koster2012snakemake} pipeline.
We aligned each sequence to the Wuhan-Hu-1 reference sequence (MN908947.3) using 
MAFFT v7.475 \cite{Katoh2013},
with the option \texttt{--keeplength} to retain the reference coordinate system.
We then used \texttt{sc2ts import-alignments} and 
\texttt{sc2ts import-metadata} to convert all alignments and metadata to VCF
Zarr format, resulting in a single 401 MiB file 
(around 77X smaller than gzip compressed FASTA\cite{Czech2025}).

A classical tradeoff in bioinformatics is whether to store data 
in sample-major (FASTA) or variant-major (VCF) format, determining
whether the data can be accessed efficiently either by row or column
of the variant matrix. Because VCF Zarr is chunked in two dimensions,
we can now access either sample aligments or the variation
data for a given site across all samples efficiently.
See Czech et al.\cite{Czech2025} for more details and benchmarks on 
this dataset. 
Another advantage of Zarr over formats such as VCF is that is
is inherently ``appendable'', and large datasets can be updated
in real time

Both the pipeline and converted Viridian dataset are available for download;
see Data and code availability for details.

@@

@@

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants,
which can be caused by correlated bases
that are treated independently by the HMM.
Correlated changes can be caused by
systematic errors in genome sequences
associated with sequencing protocols \cite{Turakhia2020},
multi-nucleotide substitutions \cite{DeMaio2024}, and
alignment errors (e.g., deletions due to MAFFT misalignment \cite{Hunt2024}).

We characterized artifactual recombination events
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
We treated adjacent and near-adjacent sites
(separated by at most three bases) as single loci.
Also, when counting loci supporting a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
Evidence that this filter effectively flagged artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events
that we identify as artifactual.
This primer scheme was used to sequence XX of the samples
associated with an artifactual recombination event,
in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
strongly implicating it as a common source of problematic data.

Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
AmpliSeq
True MNS
Alignment error

@@

This computation is a good example of type of integrated large-scale data analysis 
that can be easily performed using the combination of tskit and VCF Zarr.
The entire analysis is performed in a short Jupyter
notebook~\cite{kluyver2016jupyter}, with the 
core calculation requiring about 30 lines of Python code. Retrieving 
data from the base alignments and ARGs for 2.48 million samples 
at the same time over all 27,507 sites took only 6 minutes, with a peak memory
usage of about 5GiB of RAM on a standard desktop computer.
